Revealing Transient Strain in Geodetic Data with Gaussian Process Regression

### 2017-06-19 HinesHetlandGPR.tex
more crap
added more

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard. Knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically. Consequently, secular crustal strain rates estimated from GNSS data have been used to constrain seismic hazard models such as UCERF3 \citep{Field2014}. Dense networks of continuous GNSS stations, such as the Plate Boundary Observatory (PBO), make it feasible to estimate transient crustal strain, which we consider to be any deviation from the secular strain rate.  Transient crustal strain can potentially illuminate geophysical signal that may not be immediately apparent from the GNSS displacement data. The ability to detect transient geophysical phenomena is also relevant for seismic hazard because there are instances of major earthquakes being triggered by slow slip events \citep{Roeloffs2006} and postseismic deformation \citep{Freed2001}.

Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest and traditional method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant spatial gradients) over some subnetwork of the GNSS stations \citep[e.g.,][]{Frank1966,Prescott1976,Savage1986,Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined through a least squares fit to the observations. The assumption that deformation gradients are spatially uniform is not appropriate when subnetworks span a large area.  To help overcome this deficiency, \citet{Shen1996,Shen2015} used an inverse distance weighting scheme, in which the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996} and \citet{Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized GNSS derived velocities with bi-cubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The type of basis functions and the number of degrees of freedom for a parameterization can be subjective. If there are too few degrees of freedom in the parameterization, then estimated strain rates will be biased and the uncertainties will be underestimated. On the other hand, if there are too many degrees of freedom, then there will not be any coherent features in the estimated strain rates. The methods described by \citet{Beavan2001}, \citet{Tape2009}, and \citet{Sandwell2016} also require the user to specify penalty parameters that control a similar trade-off between bias and variance in the solution. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and fault geometries are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

The aforementioned studies are concerned with estimating secular strain rates. In recent years the Southern California Earthquake Center (SCEC) community has shown interest in developing methods for detecting transient strain. SCEC supported a transient detection exercise \citep{Lohman2013}, where several research groups tested their methods for detecting transient geophysical signals with a synthetic GNSS dataset. Among the methods tested were the Network Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion Filter (NIF) \citep{Segall1997}. the NSF uses a wavelet parameterization to  describe the spatial component of geophysical signal and the NIF, which is intended for imaging slow fault slip from geodetic data, uses the elastic dislocation Green's functions from \citet{Okada1992}. For the NSF and NIF, the time dependence of the geophysical signal is modeled as integrated Brownian motion. The method described in \citet{Holt2013} was also tested in the SCEC transient detection exercise, which calculates strain rates using a bi-cubic spatial parameterization of displacements between time epochs. \citet{Holt2013} defined a detection threshold based on the strain rate magnitude, and below we demonstrate that this is indeed an effective criterion for identifying geophysical signal. For the same reasons descibed above, the transient deformation and corresponding uncertainties estimated by these methods can be biased by the chosen spatial parameterization. It is then difficult to distinguish signal from noise with these methods, which limits their utility for transient detection.

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to infer transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Kato1998}, and it has recently gained popularity in the machine learning community. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed, continuous, stochastic process that is fully defined in terms of a mean function and a positive-definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to choose a prior Gaussian process which is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. Here we use GPR to infer transient strain resulting from slow slip events (SSEs) in Cascadia. Our results demonstrate that GPR can be an effective tool for detecting SSEs and revealing the spatio-temporal evolution of crustal strain during SSEs.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rates, which we consider to be any deviation from the secular strain rates. Our attention is limited to horizontal strain rates in this study. We denote transient crustal strain rates as $\dot{\du{\varepsilon}}(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We determine $\dot{\du{\varepsilon}}(p)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$.  As we discuss in Section \ref{sec:NoiseModel} and also as argued for in XXX, estimates of $\dot{\du{\varepsilon}}$ are more effective at illuminating geophysical signal than estimates of $\vec{u}(p)$. We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(p) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We drop the component subscripts with the understanding that the same analysis is being repeated to estimate the easting and northing components of $\vec{u}$. We assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((\vec{x},t),(\vec{x}',t')\right) = X(\vec{x},\vec{x}')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in Section \ref{sec:SignalModel}.

@@

The prior for transient displacements is then conditioned with $\bm{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that an appropriate covariance function and corresponding hyperparameters for $X$, $T$, and $C_\eta$ have already been chosen. We discuss how the covariance functions are chosen for our application to Cascadia in Sections \ref{sec:NoiseModel} and \ref{sec:SignalModel}.  If $\kappa$ is kept finite then, following \citet{Rasmussen2006}, we find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(p) = C_u(p,\bm{P})\left(\bm{\Sigma} + \kappa^2\bm{G}\bm{G}^T\right)^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(p,p') = C_u(p,p') - C_u(p,\bm{P})\left(\bm{\Sigma} + \kappa^2\bm{G}\bm{G}^T\right)^{-1}C_u(\bm{P},p').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\bm{P}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{G} \\
                         \bm{G}^T  & -\kappa^{-2} \bm{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\bm{P}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{G} \\
                          \bm{G}^T  & -\kappa^{-2} \bm{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{P},p') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\bm{P}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{G} \\
                         \bm{G}^T  & \bm{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\bm{P}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{G} \\
                          \bm{G}^T  & \bm{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{P},p') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}

We use eq. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to find the posterior easting and northing components of transient displacements. The posterior transient displacements are differentiated temporally and spatially to form an estimate of $\dot{\du{\varepsilon}}$. Using $\hat{u}_i$ to denote the posterior transient displacements along direction $i$ and $x_i$ to represent the components of $\vec{x}$, we can write the components of $\dot{\du{\varepsilon}}$ as 
\begin{equation}\label{eq:StrainRate}
\dot{\varepsilon}_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
The transient strain rate components are Gaussian processes with mean functions
\begin{equation}\label{eq:StrainMean}
\mu_{\dot{\varepsilon}_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and covariance functions
\begin{equation}\label{eq:StrainCov}
C_{\dot{\varepsilon}_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right).
\end{equation}

Our motivation for estimating transient strain rates is, in part, to detect geophysical phenomena. In Section \ref{sec:Results}, we show that strain resulting from SSEs is immediately visible from simply inspecting $\dot{\du{\varepsilon}}$. However, if our goal is to automatically detect geophysical events such as SSEs, then we need to define a signal-to-noise ratio, SNR, based on $\dot{\du{\varepsilon}}$. We consider the Frobenius norm of $\dot{\du{\varepsilon}}$, $||\dot{\du{\varepsilon}}||_\mathrm{F}$, which is also referred to as the second invariant of strain rate in the geodetic literature. We use $||\dot{\du{\varepsilon}}||_\mathrm{F}$ as a metric for the strain rate ``magnitude''. Noting that $||\dot{\du{\varepsilon}}||_\mathrm{F}$ is a random variable, we choose SNR to be the ratio of the estimated mean and standard deviation of $||\dot{\du{\varepsilon}}||_\mathrm{F}$. Using nonlinear uncertainty propagation, we find SNR to be  
\begin{equation}\label{eq:SNR}
\mathrm{SNR}(p) = \frac{\mu_{\dot{\varepsilon}_\mathrm{nn}}(p)^2 +
                        \mu_{\dot{\varepsilon}_\mathrm{ee}}(p)^2 +
                        2\mu_{\dot{\varepsilon}_\mathrm{en}}(p)^2}
                       {\big(C_{\dot{\varepsilon}_\mathrm{nn}}(p,p)\mu_{\dot{\varepsilon}_\mathrm{nn}}(p)^2 + 
                              C_{\dot{\varepsilon}_\mathrm{ee}}(p,p)\mu_{\dot{\varepsilon}_\mathrm{ee}}(p)^2 + 
                              4C_{\dot{\varepsilon}_\mathrm{en}}(p,p)\mu_{\dot{\varepsilon}_\mathrm{en}}(p)^2
                        \big)^{\frac{1}{2}}}
,
\end{equation}
where the subscripts ``n'' and ``e'' denote north and east, respectively. For simplicity, we have ignored covariances between the strain rate components in eq. (\ref{eq:SNR}), even though they are non-zero.

@@

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}. We calculate the residuals between the observations and a best fitting model, and data with residuals that are anomalously large are identified as outliers. We treat $\bm{d}_*$ as a sample of $\bm{d}$ and assume that there is no correlated noise (i.e., $\eta(p) = 0$).  The best fitting model for $\bm{d}_*$ is considered to be the expected value of the random vector $u(\bm{P}) + \bm{G}\bm{a}$ after conditioning it with non-outlier observations.  We still consider $u$ to have a separable covariance function as in eq. (\ref{eq:TransientCovariance}), and the choice for $X$ and $T$ does not need to be the same as that used in Section \ref{sec:Method}. Since outliers are determined based on how well a spatially and temporally dependent model fits the data, we are able to identify anomalous observations which may not be immediately apparent based on inspection of individual time series.

@@

We use publicly available continuous GNSS data from the University Navstar Consortium (UNAVCO) \citep{Herring2016}, which can be found at www.unavco.org. We limit the dataset to the stations and time ranges which are pertinent to seven of the most recent SSEs in the Puget Sound region. The earliest SSE considered began in August 2010, and the most recent SSE began in February 2017. We consider these most recent SSEs because the station coverage is sufficiently dense for us to use maximum likelihood methods to constrain prior models.  The positions of GNSS stations used to estimate transient strain rates are shown in Figure \ref{fig:Context}.

\begin{figure*}
\includegraphics{figures/context_map/context-map.pdf}
\caption{Positions of continuous GNSS stations used to estimate transient strain rates. The colored regions indicates the density of detected seismic tremor as determined by \citet{Wech2010}. The red dots show the positions of GNSS stations mentioned in this paper. The blue dot indicates the location of the strain rates shown in Figure \ref{fig:StrainTs} and \ref{fig:StrainMag}. The blue dashed circle demarcates the spatial extent of the tremors shown in Figure \ref{fig:StrainMag}}.    
\label{fig:Context}
\end{figure*}

>>>>

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard. Knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically. Consequently, secular crustal strain rates estimated from GNSS data have been used to constrain seismic hazard models such as UCERF3 \citep{Field2014}. Dense networks of continuous GNSS stations, such as the Plate Boundary Observatory (PBO), make it feasible to estimate transient crustal strain, which we consider to be any deviation from the secular strain rate.  Transient crustal strain can potentially illuminate geophysical signal that may not be immediately apparent from the GNSS displacement data. The ability to detect transient geophysical phenomena is also relevant for assessing seismic hazard because there are instances of major earthquakes being triggered by slow slip events \citep{Roeloffs2006} and postseismic deformation \citep{Freed2001}.

Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant deformation gradients) over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined through a least squares fit to the observations. The assumption that deformation gradients are spatially uniform is not appropriate when subnetworks span a large area.  To help overcome this deficiency, \citet{Shen1996,Shen2015} used an inverse distance weighting scheme, in which the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996,Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized GNSS derived velocities with bi-cubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The type of basis functions and the number of degrees of freedom for a parameterization can be subjective. If there are too few degrees of freedom in the parameterization, then estimated strain rates will be biased and the uncertainties will be underestimated. On the other hand, if there are too many degrees of freedom, then there will not be any coherent features in the estimated strain rates. The methods described by \citet{Beavan2001}, \citet{Tape2009}, and \citet{Sandwell2016} also require the user to specify penalty parameters that control a similar trade-off between bias and variance in the solution. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and fault geometries are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

The aforementioned studies are concerned with estimating secular strain rates. In recent years the Southern California Earthquake Center (SCEC) community has shown interest in developing methods for detecting transient strain. SCEC supported a transient detection exercise \citep{Lohman2013}, where several research groups tested their methods for detecting transient geophysical signals with a synthetic GNSS dataset. Among the methods tested were the Network Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet parameterization to  describe the spatial component of geophysical signal. The NIF, which is intended for imaging slow fault slip from geodetic data, uses the elastic dislocation Green's functions from \citet{Okada1992}. For the NSF and NIF, the time dependence of the geophysical signal is modeled as integrated Brownian motion. The method described in \citet{Holt2013} was also tested in the SCEC transient detection exercise, which calculates strain rates using a bi-cubic spatial parameterization of displacements between time epochs. \citet{Holt2013} defined a detection threshold based on the strain rate magnitude, and below we demonstrate that this is indeed an effective criterion for identifying geophysical signal. For the same reasons descibed above, the transient deformation and corresponding uncertainties estimated by these methods can be biased by the chosen spatial parameterization. It is then difficult to distinguish signal from noise with these methods, which limits their utility for transient detection.

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Kato1998}, and it has recently gained popularity in the machine learning community. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed, continuous, stochastic process that is fully defined in terms of a mean function and a positive-definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. Here we use GPR to infer transient strain resulting from slow slip events (SSEs) in Cascadia. Our results demonstrate that GPR can be an effective tool for detecting SSEs and revealing the spatio-temporal evolution of crustal strain during SSEs.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rates. Our attention is limited to horizontal strain rates in this study. We denote transient crustal strain rates as $\dot{\du{\varepsilon}}(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We determine $\dot{\du{\varepsilon}}(p)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$.  As we discuss in Section \ref{sec:NoiseModel} and also as argued for in XXX, estimates of $\dot{\du{\varepsilon}}$ are more effective at illuminating geophysical signal than estimates of $\vec{u}(p)$. We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(p) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We drop the component subscripts with the understanding that the same analysis is being repeated to estimate the easting and northing components of $\vec{u}$. We assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((\vec{x},t),(\vec{x}',t')\right) = X(\vec{x},\vec{x}')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in Section \ref{sec:SignalModel}.

@@

The prior for transient displacements is then conditioned with $\bm{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_\eta$ have already been chosen. In Section \ref{sec:NoiseModel} and \ref{sec:SignalModel}, we discuss how the covariance functions are chosen for our application to GNSS data from Cascadia. If $\kappa$ is kept finite then, following \citet{Rasmussen2006}, we find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(p) = C_u(p,\bm{P})\left(\bm{\Sigma} + \kappa^2\bm{G}\bm{G}^T\right)^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(p,p') = C_u(p,p') - C_u(p,\bm{P})\left(\bm{\Sigma} + \kappa^2\bm{G}\bm{G}^T\right)^{-1}C_u(\bm{P},p').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\bm{P}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{G} \\
                         \bm{G}^T  & -\kappa^{-2} \bm{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\bm{P}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{G} \\
                          \bm{G}^T  & -\kappa^{-2} \bm{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{P},p') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\bm{P}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{G} \\
                         \bm{G}^T  & \bm{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\bm{P}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{G} \\
                          \bm{G}^T  & \bm{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{P},p') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}

We use eq. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to find the posterior easting and northing components of transient displacements. Using $\hat{u}_i$ to denote the posterior transient displacements along direction $i$ and $x_i$ to represent the components of $\vec{x}$, we can write the components of $\dot{\du{\varepsilon}}$ as 
\begin{equation}\label{eq:StrainRate}
\dot{\varepsilon}_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
The transient strain rate components are Gaussian processes with mean functions
\begin{equation}\label{eq:StrainMean}
\mu_{\dot{\varepsilon}_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and covariance functions
\begin{equation}\label{eq:StrainCov}
C_{\dot{\varepsilon}_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right).
\end{equation}

Our motivation for estimating transient strain rates is, in part, to detect geophysical phenomena. We propose using a signal-to-noise ratio, SNR, based on $\dot{\du{\varepsilon}}$ as our detection criterion. The Frobenius norm of $\dot{\du{\varepsilon}}$, $||\dot{\du{\varepsilon}}||_\mathrm{F}$, which is also referred to as the second invariant of strain rate in the geodetic literature, is often used as a metric for the strain rate ``magnitude''. Noting that $||\dot{\du{\varepsilon}}||_\mathrm{F}$ is a random variable, we choose SNR to be the ratio of the estimated mean and standard deviation of $||\dot{\du{\varepsilon}}||_\mathrm{F}$. Using nonlinear uncertainty propagation, we find SNR to be  
\begin{equation}\label{eq:SNR}
\mathrm{SNR}(p) = \frac{\mu_{\dot{\varepsilon}_\mathrm{nn}}(p)^2 +
                        \mu_{\dot{\varepsilon}_\mathrm{ee}}(p)^2 +
                        2\mu_{\dot{\varepsilon}_\mathrm{en}}(p)^2}
                       {\big(C_{\dot{\varepsilon}_\mathrm{nn}}(p,p)\mu_{\dot{\varepsilon}_\mathrm{nn}}(p)^2 + 
                              C_{\dot{\varepsilon}_\mathrm{ee}}(p,p)\mu_{\dot{\varepsilon}_\mathrm{ee}}(p)^2 + 
                              4C_{\dot{\varepsilon}_\mathrm{en}}(p,p)\mu_{\dot{\varepsilon}_\mathrm{en}}(p)^2
                        \big)^{\frac{1}{2}}}
,
\end{equation}
where the subscripts ``n'' and ``e'' denote north and east, respectively. For simplicity, we have ignored covariances between the strain rate components in eq. (\ref{eq:SNR}), even though they are non-zero.

@@

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}. We calculate the residuals between the observations and a best fitting model. Data with residuals that are anomalously large are then identified as outliers. We treat $\bm{d}_*$ as a sample of $\bm{d}$ and assume that there is no correlated noise (i.e., $\eta(p) = 0$).  The best fitting model for $\bm{d}_*$ is considered to be the expected value of the random vector $u(\bm{P}) + \bm{G}\bm{a}$ after conditioning it with non-outlier observations.  We still consider $u$ to have a separable covariance function as in eq. (\ref{eq:TransientCovariance}), and the choice for $X$ and $T$ does not need to be the same as that used in Section \ref{sec:Method}. Since outliers are determined based on how well a spatially and temporally dependent model fits the data, we are able to identify anomalous observations which may not be immediately apparent based on inspection of individual time series.

@@

We use publicly available continuous GNSS data from the University Navstar Consortium (UNAVCO) \citep{Herring2016}, which can be found at www.unavco.org. We limit the dataset to the stations and time ranges which are pertinent to the seven most recent SSEs in the Puget Sound region. The earliest SSE considered in this study began in August 2010, and the most recent SSE began in February 2017. We use these most recent SSEs because the station coverage is sufficiently dense for us to use maximum likelihood methods to constrain prior models.  The positions of GNSS stations used to estimate transient strain rates are shown in Figure \ref{fig:Context}.

\begin{figure*}
\includegraphics{figures/context_map/context-map.pdf}
\caption{Positions of continuous GNSS stations used to estimate transient strain rates. The colored regions indicate the distribution of seismic tremor as determined by \citet{Wech2010}. The red dots show the positions of GNSS stations mentioned in this paper. The blue dot indicates the location of the strain rates shown in Figure \ref{fig:StrainTs} and \ref{fig:StrainMag}. The blue dashed circle demarcates the spatial extent of the tremors shown in Figure \ref{fig:StrainMag}.}    
\label{fig:Context}
\end{figure*}

\subsection{Noise model}\label{sec:NoiseModel}
Before we determine the transient strain rates, we must establish a prior for the transient displacements, $u$, and the noise, $\eta$. In this section we discuss our choice for the noise covariance function $C_\eta$. There have been numerous studies on temporally correlated noise in GNSS data \citep[e.g.,][]{Zhang1997,Mao1999,Williams2004,Langbein2008}. In these studies, temporally correlated noise was described with some combination of Brownian motion, a first-order Gauss-Markov (FOGM) process, and/or flicker noise. There is some physical justification for using Brownian motion as a noise model because it accurately describes the power spectrum of motion resulting from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,Wyatt1989}. Here we describe the time dependence of $\eta$ as a FOGM process and consider $\eta$ to be spatially uncorrelated. A FOGM process is a solution to the stochastic differential equation
\begin{equation}\label{eq:FOGMdef}
\dot{\eta}(t) + \alpha \eta(t) = \beta w(t),
\end{equation}
where $w(t)$ is white noise with unit variance. The FOGM process degenerates to the commonly used Brownian motion noise model under the condition that $\alpha=0$ and $\eta(0) = 0$. Our noise model that satisfies eq. (\ref{eq:FOGMdef}) is a Gaussian process with zero mean and the covariance function
\begin{equation}\label{eq:FOGM}
C_\eta\left((\vec{x},t),(\vec{x}',t')\right) = \frac{\beta^2}{2\alpha}\exp\left(-\alpha|t - t'|\right) \delta(||\vec{x} - \vec{x}'||_2). 
\end{equation}
